

Does breed influence rehoming time in shelter dogs

Step 1) Collect the data

```{r}
#Started by loading the data
load("/Users/sarahgreen/Desktop/R/rehoming.RData")
#Created unique data sample
createsample(201358111)
save(mysample, file = "mysample.RData")
#View the data
View(mysample)
table(mysample$Breed)
```

My breeds are 94 rottweilers, 84 border collies, and 12 bichon frisé.

Step 2) Cleaning the data of missing values (NA and 99999)

```{r}
rehoming_rows_removed <- nrow(mysample[mysample$Rehomed ==99999,])
rehoming_rows_removed
percentage_of_rehoming_removed <- ((rehoming_rows_removed/nrow(mysample)*100))
percentage_of_rehoming_removed
breed_rows_removed <- sum(is.na(mysample$Breed))
breed_rows_removed
percentage_of_breed_removed <- ((breed_rows_removed/nrow(mysample)*100))
percentage_of_breed_removed
total_rows_removed <- rehoming_rows_removed+breed_rows_removed
total_rows_removed
percentage_total_removed <- ((total_rows_removed/nrow(mysample))*100)
percentage_total_removed
```
Get final cleaned data

```{r}
final_clean_data <- mysample[mysample$Rehomed !=99999 & !is.na(mysample$Breed),]
View(final_clean_data)
#Splitting the data by breed for easy use
breed_data <- split(final_clean_data, final_clean_data$Breed)
breed1 <- breed_data[["Rottweiler"]]
breed2 <- breed_data[["Border Collie"]]
breed3 <- breed_data[["Bichon Frise"]]
summary(breed1)
summary(breed2)
summary(breed3)
```

Step 3) Exploring the data.
Create a table of all the mean values per breed

```{r}
mean_table <- aggregate(cbind(Visited,Rehomed,Health) ~ Breed, data = final_clean_data, FUN = mean)
colnames(mean_table) <- c("Breed", "Visited (Mean)", "Rehomed (Mean)", "Health (Mean)")
mean_table
```
Assigning a specific colour to the seperate breeds

```{r}
breed_colours <- c("lightblue", "lightgreen", "coral")
```

Creating a boxplot summarising the table.

```{r}
boxplot(Rehomed ~ Breed, data = final_clean_data,
        main = "Rehoming time by breed",
        xlab = "Breed",
        ylab = "Rehomed time (Weeks)",
        col = breed_colours)

```

Creating a Cumulative distribution graph that compares the 3 breeds by rehoming time.

```{r}
plot(ecdf(breed1$Rehomed),
     main = "Rehoming time by breed (ECDF)",
     xlab = "Rehomed time (Weeks)",
     ylab = "Cumulative distribution function (CDF)",
     col = "red",
     xlim = range(final_clean_data$Rehomed))
#adding the other lines 
lines(ecdf(breed2$Rehomed),col = "green",)
lines(ecdf(breed3$Rehomed),col = "blue",)
```

Basic normal QQ plots for all breeds 

```{r}
par (mfrow=c(1,3))

#Rottweiler
qqnorm(breed1$Rehomed,
       main = "QQ Plot for rottweiler rehoming time",
       cex.main = 1.2,
       xlab = "Theoretical quantiles",
       ylab = "Sample quantiles",
       ylim = c(0, 50))
qqline(breed1$Rehomed,
       col = "red")
#Border collie
qqnorm(breed2$Rehomed,
       main = "QQ Plot for border collie rehoming time",
       cex.main = 1.2,
       xlab = "Theoretical quantiles",
       ylab = "Sample quantiles",
       ylim = c(0, 50))
qqline(breed2$Rehomed,
       col = "green")
#Bichon frise
qqnorm(breed3$Rehomed,
       main = "QQ Plot for bichon frisé rehoming time",
       cex.main = 1.2,
       xlab = "Theoretical quantiles",
       ylab = "Sample quantiles",
       ylim = c(0, 50),
       xlim = c(-2,2))
qqline(breed3$Rehomed,
       col = "blue")

```

Adjusted QQ plots for best fit distributions and transformations.

```{r}
par (mfrow=c(1,3))
#Rottweiler
#obtain gamma data: mean, variance, shape and rate
mean_gamma <- mean(breed1$Rehomed)
variance_gamma <- var(breed1$Rehomed)
shape_gamma <- mean_gamma^2/variance_gamma
rate_gamma <- mean_gamma/variance_gamma

observed <- sort(breed1$Rehomed)
theoretical <- qgamma(ppoints(length(observed)),shape = shape_gamma, rate = rate_gamma)

plot(theoretical, observed,
      main = "Rottweiler QQ plot (Gamma)",
      xlab = "Theoretical quantiles",
      ylab = "Sample quantiles")
abline(0,1, col = "red")

#Border collie
#Square root the data to reduce the sqew.
breed2_sqrt <- sqrt(breed2$Rehomed)

qqnorm(breed2_sqrt,
       main = "Sqrt border collie QQ plot (Normal)",
       xlab = "Theoretical quantiles",
       ylab = "Sample quantiles")
qqline(breed2_sqrt,
       col = "green")

#Bichon frise
#square root of the data
breed3_sqrt <- sqrt(breed3$Rehomed)

qqnorm(breed3_sqrt,
       main = "Sqrt bichon frise QQ plot (Normal)",
       xlab = "Theoretical quantiles",
       ylab = "Sample quantiles")
qqline(breed3_sqrt,
       col = "blue")
```

Basic histograms of breed rehoming times, default breaks 

```{r}
par(mfrow=c(1,3))
#Rottweiler
hist(breed1$Rehomed,
     main = "Rottweiler",
     xlab = "Rehomed time (Weeks)",
     ylab = "Frequency",
     col = "coral")
#Border collie
hist(breed2$Rehomed,
     main = "Border collie",
     xlab = "Rehomed time (Weeks)",
     ylab = "Frequency",
     col = "lightgreen")
#Bichon frise
hist(breed3$Rehomed,
     main = "Bichon frise",
     xlab = "Rehomed time (Weeks)",
     ylab = "Frequency",
     col = "lightblue")
```

Adjusted histograms for best distributions and transformations.(default breaks)

```{r}
#install mass package
install.packages("MASS")
library(MASS)

par(mfrow=c(1,3))
#Rottweiler
hist(breed1$Rehomed,
     main = "Rottweiler",
     xlab = "Rehomed time (Weeks)",
     ylab = "Density",
     col = "coral",
     freq = F,
     ylim = c(0,0.10))
#Add the kernel density estimate line for comparison
lines(density(breed1$Rehomed), col = "darkred", lwd = 2)
#Add gamma distribution line 
curve(dgamma(x, shape_gamma, rate=rate_gamma),
      col = "chartreuse",
      add = T,
      lwd = 2)
legend ("topright",
        legend = c("Kernel density", "Gamma distribution"),
        col = c("darkred","chartreuse"),
        lwd = 2)

text (x = max (breed1$Rehomed)*0.7, y=0.08,
      labels = bquote("α = 2.86,
β = 0.27"))
#Border collie
hist(breed2_sqrt,
     main = "Border collie (Sqrt)",
     xlab = "Sqrt rehomed time (Weeks)",
     ylab = "Density",
     col = "lightgreen",
     freq = F,
     ylim = c(0,0.45))
#Add the kernel density estimate line for comparison
lines(density(breed2_sqrt), col = "darkgreen", lwd = 2)
#Add normal distribution line
normal_line <- fitdistr(breed2_sqrt, "normal")
curve(dnorm(x, mean = normal_line$estimate[1], sd=normal_line$estimate[2]),
      col = "purple",
      add = T,
      lwd = 2)
legend ("topright",
        legend = c("Kernel density", "Normal distribution"),
        col = c("darkgreen","purple"),
        lwd = 2)

text (x = max (breed2_sqrt)*0.7, y=0.36,
      labels = bquote("μ = 4.29,
σ2 = 2.07"))
#Bichon frise
hist(breed3$Rehomed,
     main = "Bichon frisé",
     xlab = "Rehomed time (Weeks)",
     ylab = "Density",
     col = "lightblue",
     freq = F,
     ylim = c(0,0.10))
#Add the kernel density estimate line for comparison
lines(density(breed3$Rehomed), col = "darkblue", lwd = 2)
#Add normal distribution line
normal_line2 <- fitdistr(breed3$Rehomed, "normal")
curve(dnorm(x, mean = normal_line2$estimate[1], sd=normal_line2$estimate[2]),
      col = "purple",
      add = T,
      lwd = 2)
legend ("topright",
        legend = c("Kernel density", "Normal distribution"),
        col = c("darkblue","purple"),
        lwd = 2)

text (x = max(breed3$Rehomed)*0.7, y=0.08,
      labels = bquote("μ = 23.40,
σ2 = 146.17"))
```
I now need the parameters of the proposed distributions
Methods of moments line could be plotted 
```{r}
#Rottweilers
mean_gamma
variance_gamma
shape_gamma 
rate_gamma
#Border collies
mean_breed2 <- mean(breed2_sqrt)
sd_breed2 <-sd(breed2_sqrt)
mean_breed2
sd_breed2
#Bichon frise
mean_breed3 <- mean(breed3$Rehomed)
sd_breed3 <-sd(breed3$Rehomed)
mean_breed3
sd_breed3
#Creating a table of all the parameters

parameter_table <- data.frame(
  breed = c("Rottweiler", "Border collie (Sqrt)", "Bichon frise"),
  Model = c("Gamma", "Normal", "Normal"),
  Parameters = c(
    sprintf("Shape: %.2f, Rate: %.2f", shape_gamma, rate_gamma),
    sprintf("Mean: %.2f, SD: %.2f", mean_breed2, sd_breed2),
    sprintf("Mean: %.2f, SD: %.2f", mean_breed3, sd_breed3)
))
parameter_table

```

Test breeds using Shapiro wilk and Kolmogorov-smirnov test.

```{r}
#Kolmogorov-smirnov test for gamma
breed1_jittered <- breed1$Rehomed + runif(length(breed1$Rehomed), -0.001, 0.001)
ks.test(breed1_jittered,"pgamma", shape_gamma, rate_gamma)
#Shapiro wilk for normal
shapiro.test(breed2_sqrt)
shapiro.test(breed3_sqrt)
shapiro.test(breed2$Rehomed)
shapiro.test(breed3$Rehomed)

ks.test(breed3$Rehomed, "punif", min=min(breed3$Rehomed),max=max(breed3$Rehomed))
```

Confidence intervals (T-tests)

```{r}
#Rottweiler
breed1_test <- t.test(breed1$Rehomed, mu=27, alternative="t", conf.level=0.95)
breed1_test
#Rottweiler Z-test
sd_b1 <- sd(breed1$Rehomed)
z_score_sd <- (mean_gamma-27)/sd_b1/(sd_b1/sqrt(93))
pvalue_z <- 2*(1-pnorm(abs(z_score_sd)))
z_score_sd
pvalue_z
#This was not used due to choosing t-test
#Border collie
breed2_test <- t.test(breed2$Rehomed, mu=27, alternative="t", conf.level=0.95)
breed2_test
#Bichon frise
breed3_test <- t.test(breed3$Rehomed, mu=27, alternative="t", conf.level=0.95)
breed3_test
```
Caterpillar plots for CI

```{r}

breed1_test <- t.test(breed1$Rehomed, mu=27, alternative="t", conf.level=0.95)

z_critical <- qnorm(0.975)
se <- sd_b1/sqrt(93)
ci_low <-mean_gamma-z_critical*se
ci_up <-mean_gamma+z_critical*se
ci_low
ci_up
#not used due to choosing t-test^

breed2_test <- t.test(breed2$Rehomed, mu=27, alternative="t", conf.level=0.95)
breed3_test <- t.test(breed3$Rehomed, mu=27, alternative="t", conf.level=0.95)

effect <- c(breed1_test$estimate, breed2_test$estimate, breed3_test$estimate)
lower <- c(breed1_test$conf.int[1], breed2_test$conf.int[1], breed3_test$conf.int[1])
upper <- c(breed1_test$conf.int[2], breed2_test$conf.int[2], breed3_test$conf.int[2])
endpoint <- c("Rottweiler", "Border collie", "Bichon frisé" )
verticalpos <- seq_along(endpoint)

caterpillar_colours <- c("darkred","darkgreen","darkblue")

par(oma=c(0,0,0,0), mar=c(6,3,1,1))
plot(x=1,
     y=0,
     type="n",
     ann=F,
     xaxt="n",
     yaxt="n",
     xlab="",
     ylab=NULL,
     bty="n",
     xlim=c(0,35),
     ylim=c(0,length(endpoint)+1))
x_lines <- seq(0,35, by=1)
abline(v=x_lines, col="grey90", lty=3)
abline(v=27, col="red", lwt=2)
#Adding the points 
for(i in seq_along(verticalpos)) {
  points(effect[i],verticalpos[i],cex=1.5,pch=16,col=caterpillar_colours[i])
  lines(c(lower[i],upper[i]), c(verticalpos[i], verticalpos[i]), lwd=1.5, col=caterpillar_colours[i])
  lines(c(lower[i],lower[i]), c(verticalpos[i]+0.2, verticalpos[i]-0.2), lwd=1, col=caterpillar_colours[i])
  lines(c(upper[i],upper[i]), c(verticalpos[i]+0.2, verticalpos[i]-0.2), lwd=1, col=caterpillar_colours[i])
  
  text(x=effect[i],
       y=verticalpos[i] + 0.4,
       labels=paste("(", round(lower[i],1), ", ", round(upper[i],1), ")", sep=""), 
       cex=0.8)
}
text(x=-0.5, y=verticalpos, label=endpoint, las=1, adj=0, cex=1)
axis(side=1,at=seq(0,35, by=5))
mtext("Mean time of dogs rehomed in weeks (95% confidence intervals)",side=1,line=4,cex=1)

```


Comparing breed means via two sample T tests

```{r}
#comparing the breeds
# 1+2
test1_2 <- t.test(breed1$Rehomed,breed2$Rehomed)
test1_2
# 1+3
test1_3 <- t.test(breed1$Rehomed,breed3$Rehomed)
test1_3
# 2+3
test2_3 <- t.test(breed2$Rehomed,breed3$Rehomed)
test2_3
```

Compare confidence intervals via a forest or caterpillar plot.

```{r}

mean_dif <- c((test1_2$estimate[1]-test1_2$estimate[2]),
              (test1_3$estimate[1]-test1_3$estimate[2]),
              (test2_3$estimate[1]-test2_3$estimate[2]))
lower <- c(test1_2$conf.int[1], test1_3$conf.int[1], test2_3$conf.int[1])
upper <- c(test1_2$conf.int[2], test1_3$conf.int[2], test2_3$conf.int[2])
endpoint <- c("Rottweiler +
Border collie", "Rottweiler +
Bichon frisé", "Border collie +
Bichon frisé" )
verticalpos <- seq_along(endpoint)

par(oma=c(0,0,0,0), mar=c(6,3,1,1))
plot(x=1,
     y=0,
     type="n",
     ann=F,
     xaxt="n",
     yaxt="n",
     xlab="",
     ylab=NULL,
     bty="n",
     xlim=c(-30,10),
     ylim=c(0,length(endpoint)+1))
x_lines <- seq(-30,10, by=1)
abline(v=x_lines, col="lightgrey", lty=3)
abline(v=0, col="grey90", lwt=2)
#Adding the points 
for(i in seq_along(verticalpos)) {
  points(mean_dif[i],verticalpos[i],cex=1.5,pch=16)
  lines(c(lower[i],upper[i]), c(verticalpos[i], verticalpos[i]), lwd=1.5)
  lines(c(lower[i],lower[i]), c(verticalpos[i]+0.2, verticalpos[i]-0.2), lwd=1)
  lines(c(upper[i],upper[i]), c(verticalpos[i]+0.2, verticalpos[i]-0.2), lwd=1)
  
  text(x=mean_dif[i],
       y=verticalpos[i] + 0.4,
       labels=paste("(", round(lower[i],1), ", ", round(upper[i],1), ")", sep=""), 
       cex=0.8)
}
text(x=-30, y=verticalpos, label=endpoint, las=1, adj=0, cex=1)
axis(side=1,at=seq(-30,10, by=5))
mtext("(Mean differences in rehoming time between breeds in weeks (95% confidence intervals)",side=1,line=4, cex=1)
```

```{r}
table_r1 <- table(breed1$Reason)
table_r1
table_r2 <- table(breed2$Reason)
table_r2
table_r3 <- table(breed3$Reason)
table_r3
```

